Decompositions of congruence subgroups of Chevalley groups

### 2015-11-07 paper.tex
Replaced biber with bibtex
Removed commented lines
Merged all sources into a single file

\usepackage[backend=biber, citestyle=numeric-comp, natbib=true, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}            
\usepackage[activate={true,nocompatibility}]{microtype}
\usepackage[capitalize]{cleveref}

@@

The purpose of this paper is two-fold. 
First, we obtain relative analogues of several classical parabolic factorizations of Chevalley groups such as the Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
Then we apply these factorizations to the study of various bounded generation properties of Chevalley groups over rings.
%To be more specific, we study relative analogues of the following three kinds of decompositions:
%\begin{itemize}
%\item \emph{Gauss decomposition, } i.\,e. a decomposition of the form $\E(\Phi, R)=\B \cdot \U^- \cdot \U$, where $\B$ denotes the standard Borel subgroup, 
%      and $\U$, $\U^-$ are the unipotent radicals of $\B$ and the opposite Borel subgroup $\B^-$;
%\item \emph{Bass---Kolster decomposition} $G = \Par_i \cdot \U_i^- \cdot \U_i \cdot \U_i^-$, where 
%      $\Par_i$, $\Par_i^-$ are the opposite maximal parabolic subgroups of $\G$ corresponding to $i$-th simple root of $\Phi$ and $\U_i$, $\U_i^-$ are their unipotent radicals;
%\item \emph{Dennis---Vaserstein decomposition} $\E(\Phi, R) = \EP_i \cdot (\U_i^- \cap \U_j^-) \cdot \EP_j$, $i\neq j$, 
%      where $\EP_i$ denotes the subgroup generated by root unipotents $x_\alpha(\xi)$ contained in $\Par_i$.
%\end{itemize}
%TODO: Insert some links to VIP's papers :)

%TODO: Illustrate each statement of the text below with a reference:
%There are two main reasons to study parabolic factorizations.
%Recall that a number of various factorizations of $\G$ are known over small-dimensional rings.
%For example, Bruhat decomposition is the key instrument in the study of linear groups over fields, whereas Gauss decomposition and unitriangular factorizations are useful tools in the study of groups over local rings.
%Moreover, for each particular analytic or arithmetic setting there is a special decomposition theorem (Iwasawa, Cartan and Iwahori decompositions, to mention a few).
%Unfortunately, neither of these decompositions can be carried over to rings of higher dimension.
%The main advantage of parabolic factorizations lies in the fact that they remain true for a broader class of rings described in terms of milder ring-theoretic assumptions formulated in terms of the \emph{stable rank} of the ground ring.
%Their another advantage is that they prove useful in the study of asymptotical properties of Chevalley groups. We are going to discuss this theme in more detail below.

@@

Let $G$ be an abstract group with a generating set $X \subset G$. 
We denote by $W(G, X)$ the \emph{width of $G$ with respect to $X$}, i.\,e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.
%The problem of finding an estimate for the width of $\G(\Phi, R)$ in terms of elementary generators $x_\alpha(\xi)$ has been extensively studied in a number of works (see e.\,g. \cite{CK83, Ka82, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}).
%Summarizing results of these works, one can hope that such width is finite only for rather narrow classes of rings of number-theoretic nature, e.\,g. localizations of Dedeking rings of arithmetic type.
%For example, W.~van der Kallen has shown that the width of $\SL(3, R)$ is infinite already for a Euclidean ring $R=\mathbb{C}[t]$.

%On the other hand, much less is known about the width of the relative group $\G(\Phi, R, I)$ and the relative elementary group $\E(\Phi, R, I)$.

@@

%\begin{rem}
%For classical groups there is a standard way to fix the signs.
%In case $\Phi=\rC_\ell$ the long and short root unipotents are
%\begin{align*}
%& x_{i, -i}(\xi)=e+\xi e_{i, -i}, \\ 
%& x_{i, j}(\xi)=e+\xi e_{ij}-\varepsilon_i\varepsilon_j\xi e_{-j, -i}, \quad i\neq\pm j.
%\end{align*}
%Here $\varepsilon_i$ is the sign of $i$.
%In case $\Phi=\rB_\ell, \rD_\ell$ the long and short root unipotents in $\SO(n, R)$ are
%\begin{align*}
%& x_{i, j}(\xi)=e+\xi e_{ij}-\xi e_{-j, -i}, \quad i\neq\pm j, \ i, j\neq0, \\
%& x_{i0}(\xi)=e+2\xi e_{i0}-\xi e_{0, i}-\xi^2 e_{i, -i}, \quad i\neq0.
%\end{align*}

@@

The following lemma is a relative version of the classical result sometimes called the Chevalley---Matsumoto decomposition.
It can be deduced easily from the absolute statement (see~\cite[Theorem~1.3]{St78}).
\begin{lemma}\label{lemma:Chevalley-Matsumoto}
Let $\pi$ be a basic fundamental representation of $\G_{sc}(\Phi, R)$ with the highest weight $\varpi_s$.
Assume that $g\in \G_{sc}(\Phi, R, I)$ is such that $(g\cdot v^+)_{\varpi\ssub{s}}=1$, then 
\[ g \in \U(\Sigma_s^-, I) \cdot \G_{sc}(\Delta_s, R, I) \cdot \U(\Sigma_s^+, I). \]
\end{lemma}
%\begin{proof}
%In the absolute case ($I=R$) the statement of the lemma is contained in~\cite[Theorem~1.3]{St78}. 
%Thus we can write $g=u_1\cdot g'\cdot u_2$, where $u_1\in\U(\Sigma_s^-, R)$, $u_2\in\U(\Sigma_s^+, R)$ and $g'\in\G(\Delta_s, R)$.
%Since $g$ lies in $\G(\Phi, R, I)$, the vector $gv^+$ is $I$-unimodular. 
%On the other hand, $g'u_2v^+=v^+$, so $u_1\in\G(\Phi, R, I)\cap\U(\Sigma_s^-, R)=\U(\Sigma_s^-, I)$.
%The matrix $g'u_2$ is an element of $\left(\G(\Delta_s, R)\U(\Sigma_s^+, R)\right)\cap\G(\Phi, R, I)$, and by Levi decomposition $u_2\in\U(\Sigma_s^+, I)$, $g'\in\G(\Delta_s, R, I)$.
%\end{proof}

@@

\section{Stability conditions}\label{sec:stability-conditions}
\input{stability-conditions}

\section{Relative parabolic factorizations} \label{sec:factorizations}
%TODO: Add references to works of Dennis, Bass, Vaserstein, Suslin etc.
In this section we formulate and prove the relative versions of the decompositions from~\cite{St78}, which will be the main technical tools throughout the next section.

\subsection{Relative Gauss decomposition}\label{sec:gauss}
\input{gauss}
\subsection{Relative Bass---Kolster decompositions}\label{sec:bass-kolster}
\input{bass-kolster}
\subsection{Relative Dennis---Vaserstein decompositions}\label{sec:dennis-vaserstein}
\input{dennis-vaserstein}

\section{Applications} \label{sec:applications}
\subsection{Bounded generation}\label{sec:boundgen}
\input{congrsubgr-boundgen}
\subsection{Subsystem factorizations}\label{sec:subsysfact}
\input{subsystem-factorizations}

>>>>

\usepackage[backend=bibtex, citestyle=numeric-comp, natbib=true, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}            
\usepackage[activate={true,nocompatibility}]{microtype}
\usepackage[capitalize]{cleveref}

@@

The purpose of this paper is two-fold. 
First, we obtain relative analogues of several classical parabolic factorizations of Chevalley groups such as the Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
Then we apply these factorizations to the study of various bounded generation properties of Chevalley groups over rings.

@@

Let $G$ be an abstract group with a generating set $X \subset G$. 
We denote by $W(G, X)$ the \emph{width of $G$ with respect to $X$}, i.\,e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.

@@

@@

The following lemma is a relative version of the classical result sometimes called the Chevalley---Matsumoto decomposition.
It can be deduced easily from the absolute statement (see~\cite[Theorem~1.3]{St78}).
\begin{lemma}\label{lemma:Chevalley-Matsumoto}
Let $\pi$ be a basic fundamental representation of $\G_{sc}(\Phi, R)$ with the highest weight $\varpi_s$.
Assume that $g\in \G_{sc}(\Phi, R, I)$ is such that $(g\cdot v^+)_{\varpi\ssub{s}}=1$, then 
\[ g \in \U(\Sigma_s^-, I) \cdot \G_{sc}(\Delta_s, R, I) \cdot \U(\Sigma_s^+, I). \]
\end{lemma}

@@

\section{Stability conditions}\label{sec:stability-conditions}
In this section we define the stability conditions used in the statements of our decomposition theorems in section~\ref{sec:factorizations}.
First, we recall the notion of the stable rank of an ideal introduced by L.~Vaserstein in~\cite{Va69, Va71}.
As we will be mainly concerned with applications to Chevalley groups, our exposition of the stable ranks is neccessarily limited to commutative rings.
The main goal of subsection~\ref{sec:rel-asr} is to define the relative version of the absolute stable rank condition, which generalizes the definition in~\cite{EO, MKV}.
Finally, in the end of this section we state several technical lemmas describing the action of certain unipotent radicals on the unimodular columns under the assumption of a suitable stability condition.

\subsection{Relative stable rank}
Recall that a row $a\in{}^n\!R$ is called \emph{$I$-unimodular} if it is congruent to $(1, 0, \ldots, 0)$ modulo $I$ and its components $a_1, a_2, \ldots, a_n$ generate $R$ as an ideal.
A column $b \in R^n$ is called $I$-unimodular if its transpose $b^t$ is an $I$-unimodular row.
We denote the set of all $I$-unimodular rows (resp. columns) by $\Umd(n, I)$ (resp. $\Ums(n, I)$).
When $I=R$, we refer to the $R$-unimodular rows and columns as simply \emph{unimodular}.

It is not hard to show that for an $I$-unimodular row $a$ there exists an $I$-unimodular column $b$ such that $ab=1$ (see~\cite[\S~2]{Va69}).

An $I$-unimodular row $a=(a_1, \ldots, a_{n+1})$ is called \emph{stable} if one can choose $b_1, \ldots, b_n\in I$ such that the row $(a_1+a_{n+1}b_1, \ldots, a_n+a_{n+1}b_n)$ is also $I$-unimodular. 
We say that $I$ satisfies the stable range condition $\SR_n(I)$ if any $I$-unimodular row of length $n+1$ is stable.
Vaserstein noted~\cite[Theorem~1]{Va71} that $\SR_n(I)$ implies $\SR_m(I)$ for any $m\geqslant n$. 
It is also clear that the condition $\SR_n(I)$ does not depend on the choice of the ambient ring $R$.
By definition, the \emph{stable rank} of $I$ (denoted $\sr(I)$) is the smallest natural number $n$ such that $\SR_n(I)$ holds (we set $\sr(I)=\infty$ if no such $n$ exists).

The following proposition summarizes the basic properties of the stable rank.
\begin{prop} \label{prop:sr_properties}
Let $R$ be an arbitrary commutative unital ring and let $I\trianglelefteq R$ be an ideal.
\begin{lemlist}
\item For any ideal $I'\subseteq I$ one has $\sr(I')\leqslant\sr(I)$, $\sr(I/I')\leqslant\sr(I)$;
\item One has $\sr\Bigl(\prod\limits_{i=1}^n R_i\Bigr)=\max\limits_{i=1}^n\bigl(\sr(R_i)\bigr)$;
\item Let $J(R)$ denote the Jacobson radical of $R$, then $\sr(R)=\sr(R/J(R))$ and $\sr(J(R))=1$;
\item Let $I\rtimes\mathbb{Z}$ denote the ring obtained by formally adjoining $1$ to $I$. Then $\sr(I\rtimes\mathbb{Z})=\max(2,\sr(I))$.
\end{lemlist}
\end{prop}
\begin{proof} See~\cite[Theorem~2.3]{Va69}, \cite{Va71}. 
\end{proof}
\begin{example}
Since the stable rank of a field equals $1$, one can conclude from the previous proposition that $\sr(R)=1$ for any semilocal ring $R$.
Other examples of rings of stable rank $1$ are provided by the Boolean rings, the ring of all algebraic integers, the ring of entire functions, the disk-algebra, etc. For this and many other examples see~\cite{VasSR1}.

%Formally adjoining a unit to a non-unital ring of stable rank $1$ produces a non-trivial example of an ideal having the stable rank strictly lower that the ambient ring.
\end{example}

\subsection{Relativization of the absolute stable rank}\label{sec:rel-asr}
For a row $a=(a_1, \ldots, a_n)\in{}^n\!R$ denote by $\mathfrak{J}(a)$ the intersection of all maximal ideals of $R$ containing $a_1, \ldots, a_n$.
It is easy to see that a row $a\in R^n$ is unimodular if and only if $\mathfrak{J}(a)=R$. 
Clearly, for any $g\in\GL(n,R)$ one has $\mathfrak{J}(a\cdot g)=\mathfrak{J}(a)$.
\begin{dfn}\label{dfn:j-stable}
We say that a row $a=(a_1, \ldots, a_{n+1})\in{}^{n+1}\!I$ can be \emph{$I$-shortened}, if there exist $c_1, \ldots, c_n\in I$ such that
$\mathfrak{J}(a_1, \ldots, a_{n+1})=\mathfrak{J}(a_1+c_1a_{n+1}, \ldots, a_n+c_na_{n+1}).$
\end{dfn}
\begin{dfn}\label{dfn:asr}
We say that an ideal $I$ satisfies the condition $\ASR_n(I)$ if it satisfies $\SR_n(I)$ and, moreover, any row $a\in{}^{n+1}\!I$ can be $I$-shortened.
\end{dfn}

It is easy to see that $\ASR_m(I)$ implies $\ASR_n(I)$ for any $n\geqslant m$. 
By definition, the \emph{absolute relative stable rank} $\asr(I)$ is the smallest natural $n$ such that $\ASR_n(I)$ holds (again we set $\asr(I)=\infty$ if no such $n$ exists).

A priori our definition of $\ASR_n(I)$ depends on $R$.
Below we will see that in fact there is no such dependence.
The following lemma is a relative version of~\cite[Lemma~8.2]{MKV}. 
\begin{lemma}\label{lemma:relative-asr-unimod}
For a commutative ring $R$ and an ideal $I \trianglelefteq R$ the following statements are equivalent:
\begin{lemlist} 
\item\label{asr-j-stable} Any row $a\in{}^{n+1}\!I$ can be $I$-shortened;
\item\label{asr-bak-like} For any $I$-unimodular row $(b, a_1, \ldots, a_n, d)\in\Umd(n+2, I)$ there exist $c_1, \ldots, c_n\in I$ 
 such that $(b+b', a_1+c_1d, \ldots, a_n+c_nd)$ is $I$-unimodular for any $b'\in J$, where $J=I \cdot a_1 + \ldots + I \cdot a_n + I \cdot d\leqslant I$.
\end{lemlist} 
\end{lemma}
\begin{proof}
Assume first that any row $a\in{}^{n+1}\!I$ can be $I$-shortened. 
In particular, for a given $I$-unimodular row $(b, a_1, \ldots, a_n, d)\in\Umd(n+2, I)$ there exist $c_1, \ldots, c_n$ such that
\[\mathfrak{J}(a_1, \ldots, a_{n+1})=\mathfrak{J}(a_1+c_1a_{n+1}, \ldots, a_n+c_na_{n+1}).\]
Therefore $(b,a_1+c_1d,\ldots,a_n+c_nd)$ is also unimodular. Of course, for any $b'\in J$ we could replace $b$ with $b+b'$ from the very start.

To show the converse take an arbitrary row $(a_1, \ldots, a_n, d)\in{}^{n+1}\!I$ and consider the $I$-unimodular row $(1, a_1, \ldots, a_n, d)\in\Umd(n+2, I).$
By the hypothesis, there exist $c_1, \ldots, c_n\in I$ such that
\[ v=(1+b', a'_1, \ldots, a'_n)=(1+b', a_1+c_1d, \ldots, a_n+c_nd) \]
is unimodular for any $b'\in J$.
Assume that there exists a maximal ideal $\mathfrak{m}\trianglelefteq R$ such that all $a'_1, \ldots, a'_n$ are contained in $\mathfrak{m}$, but at least one of the elements $d$, $a_i$ is not.
Then clearly $d\notin\mathfrak{m}$ and $I\not\subseteq \mathfrak{m}$ (otherwise $a_i=a'_i-c_id\in\mathfrak{m}$, contrary to the assumption).
Now we can find $t\in I$ such that its image $\bar{t}$ in the residue field $R/\mathfrak{m}$ equals $-\bar{1}/\bar{d}$.
This means that $1 + b' \in \mathfrak{m}$ for $b'=td\in J$ which, contradicts the unimodularity of $v$.
This shows that no such $\mathfrak{m}$ may exist and, therefore, $\mathfrak{J}(a'_1, \ldots, a'_n)=\mathfrak{J}(a_1, \ldots, a_n, d)$.
\end{proof}

Obviously, the second statement of \cref{lemma:relative-asr-unimod} does not depend on $R$, hence, as suggested by the notation, $\asr(I)$ is independent of $R$.

Let $R$ be a commutative ring. We denote by $\Max(R)$ its \emph{maximum spectrum}, i.e. the set of maximal ideals of $R$, equipped with the Zariski topology. For a topological space $X$ denote by $\dim(X)$ its usual topological dimension.
From the definition of $\asr(I)$ and \cite[Theorem~2.3]{EO} (or~\cite[Theorem~3.7]{MKV}) it follows that
\begin{equation} \label{sr-estimates} \sr(I)\leqslant\asr(I)\leqslant\asr(R)\leqslant \dim(\Max(R))+1\leqslant\dim(\Spec(R))+1. \end{equation}
There exist examples of rings for which the first inequality in the above formula is strict.
More precisely, there are rings of stable rank $1$ but of absolute stable rank $\geqslant2$, see~\cite{MKV}.
On the other hand, \cite[Theorem~1.3]{MKV} shows that $\sr(R)=\asr(R)$ for any principal ideal ring $R$.
\begin{example}
From inequality~\eqref{sr-estimates} one concludes, in particular, that 
\begin{itemize}
\item $\asr(R) = 1$ for every semilocal ring $R$;
\item $\asr(R)\leqslant2$ for every Dedekind domain $R$;
\item $\asr(A[x])\leqslant3$ for every locally principal ring $A$;
\item $\asr(F[x_1,\ldots,x_n])\leqslant n+1$ for any field $F$.
\end{itemize}
For certain classes of fields there is also a lower estimate for $\sr(F[x_1, \ldots, x_n])$ which allows to compute precisely the value of both $\sr(R)$ and $\asr(R)$.
For example, $\sr(F[x,y])=3$ for every $F$ with $\K_2^M(F)\neq0$ (see~\cite{Kru73}) and $\sr(F[x_1, \ldots, x_n])=n+1$ for $F\subseteq\mathbb{R}$, see~\cite[Theorem~8]{Va71}.
Notice also that $\sr(\mathbb{Z}[x])=3$ (see \cite{GMV}) and $\sr(\mathbb{Z}[x_1,\ldots,x_n])=n+1$ for $n>1$ (see example after~\cite[Theorem~18.2]{VS76}).
\end{example}

\subsection{Action of unipotent radicals}\label{sec:ur-action}
\begin{lemma}\label{lemma:PSV-symplectic-trick}
 Let $v=(v_1, \ldots, v_n)^t$ be a column. Denote by $v'$ the vector composed of squares of the components of $v$, i.\,e. $v'=(v_1^2, \ldots, v_n^2)^t$.
 Then for any matrix $b \in M(n, I)$ one can find a symmetric matrix $a \in M(n, I)$, $a=a^t$ such that $b \cdot v' = a \cdot v$. \end{lemma}
\begin{proof}
Straightforward computation shows that the assertion of lemma holds for the matrix $a$ defined by
\begin{equation*}
a_{ij} = b_{ij} v_{j} + b_{ji} v_{i},\ j\neq i,\quad a_{ii} = b_{ii} v_{i} - \sum\limits_{\mathclap{\substack{j=1 \\ j\neq i}}}^\ell b_{ji} v_{j}. \qedhere
\end{equation*}
\end{proof}

Let $v\in V=R^{2\ell}$ be a vector of the natural representation of $\G(\rD_\ell, R)$.
Denote by $v_+$ and $v_-$ the upper and the lower halves of $v$, i.\,e. $v_+=(v_1, \ldots, v_\ell)^t$, $v_-=(v_{-\ell}, \ldots, v_{-1})^t$.
\begin{lemma}\label{lemma:asrUnip}
For any $I$-unimodular column $v=(v_+, v_-)^t\in\Ums(2\ell, I)$ there exists $g\in\U(\Sigma^+_\ell, I) \leq \E(\Phi, R, I)$ 
such that $(g \cdot v)_+ \in \Ums(\ell, I)$ under the following assumptions on $\Phi$ and $I$.
\begin{lemlist}
\item \label{item:asrUnipC} $\Phi=\rC_\ell$ and $\sr(I) \leqslant \ell$;
\item \label{item:asrUnipD} $\Phi=\rD_\ell$ and $\asr(I)\leqslant \ell -1$.
\end{lemlist}
\end{lemma}
\begin{proof} 
\textsc{Case $\Phi=\rC_\ell$.}
Denote by $p$ the matrix of size $\ell$ such that its only nonzero entries equal $1$ and are on the skew-diagonal, i.\,e. $p_{ij}=\delta_{i, \ell-j+1}$. 
For $b \in M(\ell, I)$ set $g(b)=\left(\begin{smallmatrix} e_\ell & p \cdot b \\ 0 & e_{\ell} \end{smallmatrix}\right)$.
Clearly, if $b$ is symmetric then $g(b)$ lies in $\U(\Sigma_\ell^+, I)\leq \E(\rC_\ell, R, I)$.

Notice that the column $v'=(v_1,\ldots,v_\ell, v_{-\ell}^2,\ldots,v_{-1}^2)^t$ is $I$-unimodular.
By the definition of the relative stable rank we can find a matrix $b\in \M(\ell,I)$ such that the upper half $v''_+$ of the vector $v''= g(b) \cdot v'$ is $I$-unimodular.
It is clear that $v''_+ = v_+ + pb v'_-$. 
Finally, applying \cref{lemma:PSV-symplectic-trick}, we find a symmetric matrix $a$ such that
\[ (g(a)\cdot v)_+=v_+ + pav_- = v_+ + pbv'_- = v''_+ \in \Ums(\ell, I). \]

\textsc{Case $\Phi=\rD_\ell$.} Denote by $J$ the ideal of $R$ spanned by the components of $v_{-}$. Clearly, $J \subseteq I$.
By \cref{prop:sr_properties} $\sr(I/J) \leqslant \ell-1$, therefore
the elementary group $\E(\rA_{\ell-1}, R/J, I/J)$ acts transitively on $\Ums(\ell, I/J)$ (see~\cite[Theorem~2.3c]{Va69}). 
This implies the existence of an element $h\in \E(\Delta_\ell, R, I)$ such that the vector $v' = h \cdot v$ satisfies $v'_i \equiv \delta_{i1} \pmod J$ for $i=1, \ldots, \ell$.

Clearly, $(v'_1, v'_{-\ell}, \ldots, v'_{-1})$ is $I$-unimodular.
Applying \cref{asr-bak-like}, we find $c_2, \ldots, c_\ell\in I$ such that for $v''= \prod_{i=2}^{\ell}x_{-i, -1}(c_i)\cdot v'$ one has
$(v''_1, v''_{-\ell},\ldots, v''_{-2})\in\Ums(\ell+1, I)$.
Now, applying the condition $\sr(I) \leqslant \ell-1$ once again, we find
$d_1, d_3, \ldots, d_{\ell}\in I$ such that the entries $(v'''_1, v'''_{-\ell}, \ldots, v'''_{-3})$
of $v'''=x_{-2, 1}(d_1) \cdot \prod_{i=3}^{\ell} x_{-2, -i}(d_i) \cdot v''$ form an $I$-unimodular column.

We can find $f_1, f_3,\ldots, f_\ell \in R$ such that $f_1v'''_1+\sum_{i=3}^\ell f_i v'''_{-i} = 1$.
Set $\xi = v'''_1-v'''_2-1 \in I$, $v^{(4)}=x_{1,2}(\xi f_1) \cdot \prod_{i=3}^\ell x_{-i,2}(\xi f_i) \cdot v'''$.
Clearly $v^{(4)}_2 = v^{(4)}_1-1$, therefore $v^{(4)}_+$ is $I$-unimodular.
Summarizing the above, we have found $g\in \EP_\ell(R, I)$ such that $v^{(4)}=g \cdot v$
and the assertion of the lemma immediately follows from the Levi decomposition. \end{proof}

\begin{cor} \label{cor:embeddingBD}
 Denote by $i$ the standard embedding $\G(\rB_{\ell-1}, R, I) \hookrightarrow \G(\rD_\ell, R, I)$. 
 If $\asr(I) \leqslant \ell-1$, then for every $I$-unimodular column $v \in \Ums(2\ell, I)$ there exists $g \in \U(\Sigma_{\ell-1}^+, I) \leqslant \E(\rB_{\ell-1}, R, I)$ such that
 the upper half of the vector $i(g) \cdot v$ forms an $I$-unimodular column of height $\ell$.
\end{cor}
\begin{proof}
 The assertion follows from the fact that for every $u \in \U(\Sigma_\ell^+, I) \leqslant \E(\rD_\ell, R, I)$ one can find $u' \in \G(\Delta_\ell, R, I)$ such that $uu'$ lies in the image of $i$. 
 Indeed, denote by $\sigma$ the symmetry of the Dynkin diagram of $\rD_\ell$,
 and decompose $u$ into the product $u_1 \cdot u_2$, where $u_1 \in \U(\Sigma_\ell^+\cap\Delta_{\ell-1}, I)$ and $u_2 \in \U(\Sigma_\ell^+ \cap \Sigma_{\ell-1}^+, I)$.
 Then one takes $u' = \sigma(u_1)$.

 Now by \cref{item:asrUnipD} one finds a suitable $u \in \U(\Sigma_\ell^+, I) \leqslant \E(\rD_\ell, R, I)$ which makes the upper half unimodular,
 and since $u'$ lies in $\G(\Delta_\ell, R, I)$, the upper half remains unimodular.
\end{proof}

\begin{lemma} \label{lemma:uraction} 
Let $\Phi=\rA_\ell, \rC_\ell, \rD_\ell$. Denote by $\pi$ the natural representation of $\G(\Phi, R)$ on $V=R^n$, $n=\ell+1, 2\ell, 2\ell$ respectively.
Assume that one of the following assumptions holds:
\begin{lemlist}
 \item \label{item:uractionA} $\Phi=\rA_\ell$, $\Gamma=\{ k+1, \ldots, \ell+1\} \subset \Lambda(\pi)$ and $\sr(I)\leq k\leq \ell$;
 \item \label{item:uractionC} $\Phi=\rC_\ell$, $\Gamma=\{-\ell,\ldots, -2, -1\} \subset \Lambda(\pi)$ and $\sr(I)\leq \ell$. 
 \item \label{item:uractionD} $\Phi=\rD_\ell$, $\Gamma=\{-\ell,\ldots, -2, -1\} \subset \Lambda(\pi)$ and $\asr(I)\leq \ell-1$. 
\end{lemlist}
Then for any $g\in \G(\Phi, R, I)$ there exist $x\in \U(\Phi^+, I)$, $y\in \U(\Phi^-, I)$ such that $(yxg \cdot v^+)_\lambda = 0$ for all $\lambda\in \Gamma$.
\end{lemma}
\begin{proof} Denote by $v$ the image of the highest weight vector $v^+$ under $g$.

\textsc{Case $\Phi=\rA_\ell$.} From the definition of the relative stable rank it follows that we can find 
$x= \left(\begin{smallmatrix} e_k & a \\ 0 & e_{n-k} \end{smallmatrix}\right) \in \U(\Sigma_k^+, I)$ such that 
the upper $k$ components of $v'= x \cdot v$ form an $I$-unimodular column. 
Now, to obtain zeroes at the desired positions it remains to subtract from $v'_{k+1},\ldots, v'_{\ell+1}$ suitable multiples of $v'_1, \ldots, v'_k$.
This operation corresponds to the left multiplication by some element $y\in\U(\Sigma_k^-, I)$.

\textsc{Case $\Phi=\rC_\ell$.} 
Applying \cref{item:asrUnipC}, we find $x \in \U(\Sigma_\ell^+, I)$ such that the upper half $v'_+$ of $v' = x \cdot v$ is unimodular.
Set $g(a) = \left(\begin{smallmatrix} e_\ell & 0 \\ p \cdot a & e_{\ell} \end{smallmatrix}\right)$.
Clearly, if $a$ is symmetric, then $g(a) \in \U(\Sigma_\ell^-, I)$.
Since the column $v''_+ = ({v'_1}^2, \ldots, {v'_\ell}^2)^t$ is $I$-unimodular, there exists a matrix $b \in M(\ell, I)$ such that $v'_- + p b v''_+ = 0$.
Finally, using \cref{lemma:PSV-symplectic-trick}, we find a symmetric matrix $a$ such that $(g(a) \cdot v')_- = p a v'_+ + v'_- = p b v''_+ + v'_- = 0$.

\textsc{Case $\Phi=\rD_\ell$.} From the proof of \cref{item:asrUnipD} it follows that there exists $h_1 \in \EP_\ell(R, I)$ such that for $v'=h_1\cdot v$ one has $v'_2=v'_1-1\in I$.
Clearly, for $v'' = z_{-\alpha\ssub{2}}(-v'_2, 1)\cdot v'$ one has $v''_1=1$, hence by \cref{lemma:Chevalley-Matsumoto} there exists $h_2 \in \U(\Phi^-, I)$ such that the element
$g'=h_2 \cdot z_{-\alpha\ssub{2}}(-v'_2, 1) \cdot h_1 \cdot g$ fixes $v^+$. 
Using the Levi decomposition we can write $g'=h \cdot y \cdot x \cdot g$ for some $y\in\U(\Sigma^-_\ell, I)$, $x \in \U(\Sigma^+_\ell, I)$, $h\in\E(\Delta_\ell, R, I)$.
It is clear that $x$, $y$ are the desired elements.
\end{proof}s

\section{Relative parabolic factorizations} \label{sec:factorizations}
In this section we formulate and prove the relative versions of the decompositions from~\cite{St78}, which will be the main technical tools throughout the next section.

\subsection{Relative Gauss decomposition}\label{sec:gauss}
The proof of the Gauss decomposition presented below is similar to the absolute case (cf.~\cite[Theorem~5.1]{Sm12}).
\begin{prop}\label{thm:Gauss}
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ and let $\Delta_1$, $\Delta_\ell$ be
two reductive subsystems of $\Phi$, corresponding to the endnodes of the Dynkin diagram of $\Phi$.
Suppose that both relative elementary subgroups $\E(\Delta_i, R, I)$, $i=1, \ell$ admit the triangular decomposition with $N$ triangular factors:
\[ \E(\Delta_i, R, I) = \Hh(\Delta_i, R, I) \cdot \U(\Delta^+_i, I) \cdot \U(\Delta^-_i, I) \cdot \ldots \cdot \U(\Delta^\pm_i, I), \quad i=1, \ell. \]
Then the ambient group $\E(\Phi, R, I)$ admits the decomposition with the same number of factors:
\[ \E(\Phi, R, I) = \Hh(\Phi, R, I) \cdot \U(\Phi^+, I) \cdot \U(\Phi^-, I) \cdot \ldots \cdot \U(\Phi^\pm, I). \]
\end{prop}
\begin{proof}
Denote by $Y$ the product of subgroups in the right-hand side of the above equality.
To show that $Y=\E(\Phi, R, I)$ it suffices to check that
\begin{enumerate}
\item $Y$ is normalized by $\E(\Phi, R)$, i.\,e. $Y^{\E(\Phi, R)}\subseteq Y$;
\item there exists a set $X$, generating $\E(\Phi, R, I)$ as a \emph{normal} subgroup of $\E(\Phi, R)$ such that $XY\subseteq Y$.
\end{enumerate}
To prove the first assertion it suffices to show that $Y^{x_\alpha(\xi)} \subseteq Y$ for any $\alpha\in \pm \Pi$, $\xi\in R$.
Fix a root $\alpha\in\pm\Pi$. Clearly, $\alpha \in \Delta_i$ for $i=1$ or for $i=\ell$, and we can expand $Y$ as follows:
\[ Y=\Hh(\Phi, R, I) \cdot \U(\Delta_i^+, I) \cdot \ldots \cdot \U^\pm(\Delta_i, I) \cdot \U(\Sigma_i^+, I) \cdot \ldots \cdot \U(\Sigma_i^\pm, I). \]
For every  $h \in \Hh(\Phi, R, I)$ one has for some $s\in I$
\[ x_\alpha(\xi)\cdot h = h\cdot x_\alpha((1+s)\xi)=h\cdot x_\alpha(\xi)\, x_\alpha(s\xi). \]

Therefore by the assumption of the theorem
\begin{multline*}
Y^{x_\alpha(\xi)} \subseteq \Hh(\Phi, R, I) \cdot x_\alpha(-\xi) \cdot X_{\alpha}(I) \cdot \E(\Delta_i, R, I)\cdot x_\alpha(\xi) \cdot \U(\Sigma^+_i, I) \cdot \ldots \cdot \U(\Sigma^\pm_i, I) =\\
= \Hh(\Phi, R, I) \cdot \E(\Delta_i, R, I) \cdot \U(\Sigma^+_i, I) \cdot \ldots \cdot \U(\Sigma^\pm_i, I) = Y.
\end{multline*}

Now set $X=\left\{x_\alpha(\xi)\ \middle|\ \alpha\in\Pi, \ \xi\in I \right\}$. 
Every root is a conjugate of some fundamental root under the action of $W(\Phi)$.
Since $\E(\Phi, R)$ contains the normalizer of the torus, the set $X^{\E(\Phi, R)}$ contains all the generators of the group $\E(\Phi, I) = \langle x_\alpha(\xi), \  \alpha\in\Phi, \ \xi\in I \rangle$.
Finally, the inclusion $XY \subseteq Y$ follows from the fact that $\Hh(\Phi, R, I)$ normalizes every root subgroup $X_\alpha(I)$.
\end{proof}

\begin{proof}[Proof of \cref{thm:srRI1}]
In view of the above proposition it suffices to prove that the Gauss decomposition of length $3$ holds for $\Phi=\rA_1$.

Let $A=\begin{psmallmatrix}a & b \\ c & d\end{psmallmatrix}$ be an element of $\SL(2, R, I)$.
The first column of $A$ is $I$-unimodular, therefore there exists $z\in I$ such that $a+cz\in R^*$.
Multiplying $A$ on the left by $x_{12}(z)$, we get a matrix $A'=x_{12}(z)\cdot A=\begin{psmallmatrix}a' & b' \\ c & d\end{psmallmatrix}$ with the invertible  element $a'$ in the top-left corner.
After multiplying $A'$ on the left by $x_{21}(-c/a')$ and on the right by $x_{12}(-b'/a')$ we get a diagonal matrix. 
Thus we have obtained the relative Gauss decomposition of $A$
\begin{equation*}
A=x_{12}(-z)\cdot x_{21}(c/a')\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{12}(b'/a')=x_{12}(-z)\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{21}(y) \cdot x_{12}(b'/a'), 
\end{equation*}
where $\varepsilon\in 1+I$ and $y\in I$. \end{proof}
